---
title: "Yearly Twitter Sentiment Analysis on Climate"
author: "Daniel Cardenas"
format: html
editor: visual
params:
  year: 2009
---

# How was `r params$year` in climate sentiment compared to the previous year?

```{r}
#| echo: false
#| output: false 

library(tidyverse)
library(scales)

twitter = read.csv("./Twitter.csv")

twitter$year = as.integer(substr(twitter$created_at, 1, 4))

twitter <- twitter %>% 
  filter(year == params$year | year == params$year - 1) %>%
  mutate(year_current = year == params$year)

total_tweets = nrow(twitter)
current_tweets = nrow(filter(twitter, twitter$year==params$year))
past_tweets = nrow(filter(twitter, twitter$year==params$year-1))
past_year = params$year - 1
current_year = params$year
```

The universe of tweets for this analysis is `r comma(total_tweets)`.

Of those, `r comma(current_tweets)` are from `r current_year` and `r comma(past_tweets)` are from `r past_year`.

## Sentiment

The first analysis refers to the sentiment on the tweets regarding climate change for each year. The sentiment is how likely it is that positive (towards 1) or negative (towards -1) words were used in an average tweet.

```{r}
#| echo: false
#| output: true 

mean_sentiment_current = mean(filter(twitter, twitter$year == params$year)$sentiment)
mean_sentiment_past = mean(filter(twitter, twitter$year == params$year-1)$sentiment)

ggplot(twitter, aes(x = factor(year), y = sentiment, fill = factor(year))) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  scale_fill_brewer(palette = "Set3") + 
  labs(title = "Sentiment Analysis Between Years",
       x = "Year",
       y = "Sentiment Score") +
  theme_minimal() +
  guides(fill="none")
```

This graph presents a visual description of the distribution of the sentiment scores for the tweets between the two years. The mean for `r current_year` is `r round(mean_sentiment_current, 3)` and for the previous year, `r past_year`, the mean is `r round(mean_sentiment_past, 3)`.

## Agressiveness

The second analysis refers to the aggressiveness of the tweet. In particular, we want to see if over time the conversation has gone more aggressive regarding climate change.

```{r}
#| echo: false
#| output: true 

plot_df <- twitter %>%
  group_by(year) %>%
  mutate(total_per_year = n()) %>%
  ungroup() %>%
  group_by(year, aggressiveness) %>%
  summarise(
    count_ = n(),
    perc_ = count_ / first(total_per_year) * 100,
    .groups = 'drop'
  ) %>%
  arrange(desc(perc_)) 

sum_aggressive_current = sum(filter(plot_df, plot_df$year == params$year & plot_df$aggressiveness=="aggressive")$count_)
sum_aggressive_past = sum(filter(plot_df, plot_df$year == params$year-1 & plot_df$aggressiveness=="aggressive")$count_)

# Plot 1: Bar Plot
ggplot(plot_df, aes(x = factor(year), y = perc_, fill = aggressiveness)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = sprintf("%s\n%.2f%%", aggressiveness, perc_)), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 4) + 
  scale_fill_manual(values = c("#fb8072", "#ffffb3")) +
  labs(title = "Aggressiveness Percentages by Year",
       x = "",
       y = "Percentage",
       fill = "Aggressiveness") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip()

```

In absolute terms, in `r past_year` there were `r comma(sum_aggressive_past)` tweets classified as aggressive. In contrast, in `r current_year` there were `r comma(sum_aggressive_current)`.

## Stance

The third is the stance that each tweet defends, either if it is a believer that humans are responsible for climate change, a denier or if it is a neutral tweet.

```{r}
#| echo: false
#| output: true 


df_plot1 <- twitter %>%
  group_by(year) %>%
  mutate(total_per_year = n()) %>%
  group_by(year, stance) %>%
  summarise(
    count_ = n(),
    percentage = count_ / first(total_per_year) * 100,
    .groups = 'drop'
  )

most_common_stance_info <- df_plot1 %>%
  group_by(year) %>%
  slice_max(order_by = count_, n = 1) %>%
  ungroup() %>%
  slice_max(order_by = count_, n = 1)


ggplot(df_plot1, aes(x = factor(year), y = count_, fill = stance)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(comma(count_), "\n(", sprintf("%.1f%%", percentage), ")")),
            position = position_dodge(width = 0.9), 
            vjust = -0, 
            color = "black", 
            size = 3) + 
  scale_fill_brewer(palette = "Set3") + 
  scale_y_continuous(labels = comma) +
  labs(title = "Classification of tweets according to stance",
       x = "",
       y = "Count of Tweets",
       fill = "Stance") +
  theme_minimal() +
  theme(legend.position = "right") +
  guides(fill = guide_legend(ncol = 1)) +
  coord_cartesian(clip = "off")
```

Between the two years the category with the most tweets was **`r most_common_stance_info$stance`** in `r most_common_stance_info$year` with a total of `r comma(most_common_stance_info$count_)` tweets that were `r round(most_common_stance_info$percentage, 2)`% of the total tweets of that year.

## Summary

```{r}
#| echo: false
#| output: true 

higher_lower_sent = ifelse(mean_sentiment_current<=mean_sentiment_past, "lower", "higher")
higher_lower_aggr = ifelse(sum_aggressive_current<=sum_aggressive_past, "lower", "higher")
higher_lower_stan = ifelse(most_common_stance_info$year!=params$year, "lower", "higher")
```

We see that, for these statistics:

-   `r current_year` has **`r higher_lower_sent`** average tweet sentiment than the previous year.
-   `r current_year` has **`r higher_lower_aggr`** number of aggressive tweets than the previous year.
-   `r current_year` has **`r higher_lower_stan`** number of tweets classified as "`r most_common_stance_info$stance`" than the previous year.
